---
title: 'DATA 607: WEEK 5 Assignment Tidying and Transforming Data'
author: "James Williams"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

#STEP 1: READ IN DATA
Load libraries and import CSV file
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
flights <- as_tibble(read.csv("https://raw.githubusercontent.com/jambawilliams/DATA607WEEK5/master/Flights.csv", na.strings=c("","NA")))
flights
```

#STEP 2: TIDY DATA
Rename columns, remove rows with null values, fill empty cells with preceding values
```{r}
flights <- flights %>% rename(Airline = X)
flights <- flights %>% rename(Status = X.1)
flights <- flights %>% rename(Los_Angeles = Los.Angeles)
flights <- flights %>% rename(San_Diego = San.Diego)
flights <- flights %>% rename(San_Francisco = San.Francisco)
flights <- flights %>% filter(!is.na(flights$Los_Angeles))
flights <- flights %>% fill(Airline)
flights
```

#STEP 3: COMPARE ARRIVAL DELAYS
Transform tibble to organize data by destination and delay count to compare airline performance
```{r}
delay <- gather(flights, "Destination", "n", 3:7)
delay <- spread(delay, "Status", "n")
delay <- arrange(delay, desc(Destination))
delay <- delay %>% rename(Delayed = delayed)
delay <- delay %>% rename(On_Time = "on time")
delay
```
Compare total number of delays between airlines
```{r}
delay %>% group_by(Airline) %>% summarise(Delayed=sum(Delayed), On_time=sum(On_Time))
```
Compare number of delays between destinations
```{r}
delay %>% group_by(Destination) %>% summarise(Delayed=sum(Delayed), On_time=sum(On_Time)) 
```
Visualize airline performance
```{r}
graph <- ggplot(delay, aes(x = Destination, y = Delayed))+
  geom_line(aes(color=Airline, , group = Airline))
graph
```


#STEP 4: CONCLUSIONS
In absolute terms, American West had more delays than Alaska. American West though also had twice as many flights as Alaska.  American West actually had a lower rate of delay (12%) compared to Alaska (15%).  Phoenix had the most number of delays of any destination while San Diego had the least.  San Francisco had the highest rate of delays of any destination at 28%, whereas San Diego had the lowest at 8%. 


